% -*- root: ../main.tex -*-
\chapter{Introduction}
\section{The art of computer simulations}
The abundance of cheap computing resources developing over the last few decades has resulted in extensive computational studies in all parts of science. These studies have aided the understanding of physical phenomena, and the developent of new technologies. An early example is the atomic bomb. An everyday example of the application of computer simulations, is weather forecasting. Increasing access to high performace computers have resulted in weather forecasts that are useful several days in advance. In the first example -- the atomic bomb -- computer simulations aided physical understanding. In weather forecasting however, we already know the physics. The model incorporates known laws of nature to predict weather conditions forward in time. I will be doing the former -- but on another topic.

A physical model, in the sense that I will be using it, is a mathematical formulation of a physical problem with the purpose of capturing some aspect of the behavior of that system. A simulation is a solution of the equations defining the model. For some simple models, the equations can be solved analytically, and in that case, the solution can be thought of as a simulation. For many models, we are not able to solve the equations analytically, and numerical methods must be applied to simulate the model -- this is what is usally referred to as a simulation.

Computer simulations are superior when it comes to having the full state of a system available in all relevant timescales and spatial scales, since these are available by design. That means simulations can aid understanding and predicting the behavoir of systems on scales that are not experimentally available. At the same time, computer simulations -- and physical models -- exist in close interplay with experiments. For a simulation of a model to be valueable, it has to reproduce relevant features of the physical system. These features are usually found in experiments. When a models is sufficiently calibrated from experimental data, it is possible -- with some uncertainty -- to use that model to study features that were not available from the experiment, and to propose new experiments that can shed light on the findings from the model, and possibly increase confidence in that model.

Computer models are a new thing, in the sense that running simulations on electronic computers was impossible before such devices were available. On the other hand, computer models are an old thing, since mathematical models could be formulated for computers long before electronic computers were even proposed. Indeed, iterative schemes like Newtons method for zero-points of a function and Eulers method for the integration of differential equations were developed by people who never lived to see electronic computers. The computations would be performed by a human, who would at the time be referred to as a computer. 

In the interception between computational science and theoretical physics, we find the field of computational physics: Mathematical models are developed to answer physics questions, and these models are studied using methods from computational science. On the other hand, computational physics is a close relative of experimental physics, in the way that the computational physicist perform ``numerical experiments'' when running simulations. But, the simulations are fully decoupled from the sensory world -- they are not real experiments. Thus the computational physicist is a theoretical physicist using a computer. But still, experimental intuition remains important, as computer simulations resemble experiments, and the analysis methods are similar.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methane hydrates}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Methane hydrates are so called clathrate compunds, meaning that they consist of host molecules forming a lattice that traps guest molecules. Methane hydrates are a special case of \emph{gas hydrates}, or \emph{clathrate hydrates} which they are also called. In methane hydrates, water molecules form cages that can host single methane molecules. At first sight, a piece of methane hydrate will resemble regular ice. However, if kept at regular temperatures and pressures, it will release methane gas, allowing it to be set on fire -- fiery ice. 

When multiphase piplelines were introduced in the petroleum industry, finding the necessary physical conditions for methane hydrate precipitation became interesting and necessary. Multiphase pipelines -- pipelines that transport oil, natural gas and water at the same time -- are at risk of providing conditions at which methane hydrates can form a plug that will decrease or stop flow in the pipeline, and possibly damage equipment.

Over the last 10-20 years, rising prices on fossil fuels have started a revolution of research, exploration and recovery of unconventional hydrocarbon resources. Indeed, around 50 \% of all natural gas recovered in the United States in 2014 came from oil shales, and around half of the United States oil production was tight oil.

Methane hydrates have not gotten nearly as much attention as shale gas, despite the estimated amounts of hydrocarbon residing in methane hydrates being significantly larger than those of shale hydrocarbons. That might be beacuse methane hydrates seem harder to extract. Gas hydrate extraction is still in the piloting phase, and it is still not commercially viable.

In addition to the interest in methane hydrates as an energy resource, is also the fear that methane hydrates can -- if released -- can contribute significantly to out-of-control global warming. Methane hydrates are stable within some range of temperatures and pressures, but if brought out of that range, the dissociate, and release the methane. Methane is a powerful greenhouse gas, and the idea is that the released methane will contribute to even more warming, releasing more methane in a process spinning out of control: \emph{The clathrate hydrate gun hypothesis}. This hypothesis is set both as a possible explanation of past events of global warmin (on geological timescales), and a fear for the future.

Understanding the dissociation mechanisms of methane hydrates is important both with respect to extraction and with respect to the possibly hazardous behavior of methane hydrates during global warming. 

The goal for my master project is to study dissociation mechanisms for methane hydrates on the molecular scale. I will use molecular dynamics to study dissociation through crack propagation due to externally applied stress in pure, crystalline methane hydrates with artificial defects. The aim is to reproduce some simple mechanical and fracture properties in these systems, but not to reproduce or explain any particular experiments or phenomena. Additionally, some possibly fruitful paths for future research will be identified.

\section{The ethics of hydrocarbon research}
Methane hydrates can be studied with the sole purpose of understanding their basic behavior, offering no attention to how that understanding might be used. But the reality is that if the political climate doesn't change during the coming years, knowledge about new hydrocarbons can result in higher CO$_2$-emissions. Therefore, the ethics of petroleum research need to be adressed.

The Intergovernmental Panel on Climate Change (IPCC) concludes in its synthesis report from 2014 \cite{IPCC2014} that:

\begin{quotation}
The IPCC is now 95 percent certain that humans are the main cause of current global warming. In addition, the SYR finds that the more human activities disrupt the climate, the greater the risks of severe, pervasive and irreversible impacts for people and ecosystems, and long-lasting changes in all components of the climate system. 
\end{quotation}

This means that we cannot continue burning fossil fules at the rates that we have been doing over the last decade. A question naturally arising from that, is whether it is responsible to do research that can potentially increase CO$_2$-emissions by making it easier to extract fossil fuels.

Fossil fuel dependence must by definition end at some point. Either by replacement for someting else, or by fossil fuels turning out not to be fossil (which is highly unlikely). But fossil fuel extraction cannot stop tomorrow. If that happened, people would start dying, since energy-intensive fossil-fueled activity is essential for food production, healthcare, transportation and warming houses in cold areas -- we survive because of our standard of living.

Fossil fuels can be ranked by how much CO$_2$ they emit per Joule of energy produced. In that respect, it would for instance be better to research gas than coal. But gas is not good enough, unless it is only a bridge towards a fossil-free world.

Technologies for carbon capture and storage (CCS) are under development, and if they succeed, climate is no longer the limiting factor for using fossil fuels. Then the game is suddenly changed, and research that enables extracion of previously unviable resouces.

The Norwegian National Committee for Research Ethics in Science and Technology (NENT) conducted in 2014 an assessment of the ethics of norwegian petroleum research. Their main conclusion, as stated on their webpage is:

\begin{quotation}
It is indefensible from a research ethics perspective if petroleum research hinders processes of transition to sustainable energy and thus prevents achievement of UN climate goals which Norway has pledged to uphold
\end{quotation}

Which is not really a conclusion, since the premise is still very much up for debate. Does petroleum research hinder the transition to sustainable energy?

I end here by accepting that there are ethical concerns regarding peroleum research, but it is not obvious or concluded that it is wrong and needs to be stopped. 

\section{My contribution}
A master thesis contains descriptions of the works of other people in addition to the authors comtributions. This feature of the masters thesis calls for a detailed account of what are my scientific contributions.

I have developed some analysis tools, which are described in chapter \ref{ch:tools}. It should be clear from the description what parts of that chapter are my descriptions of tools made by others, and what tools were developed by me.

Most of my work has been on the properties of methane hydrates in the TIP4P/ICE water and OPLS-UAM methane model (explained in detail in chapter \ref{ch:models}). This model, like all molecular dynamics models of methane hydrates, is poorly investigated. Additionally, fracture of methane hydrates is poorly studied, and as far as I know, this is the first study of fracture of methane hydrates using molecular dynamics. 

Below follows a list of my scientific contributions:

\begin{itemize}
\item Transport properties of the TIP4P/ICE water model at \SI{300}{\kelvin}. (Diffusion coefficient and viscosity).
\item Mechanical properties of methane hydrates in TIP4P/ICE+OPLS+UAM: Young's modulus, Poisson's ratio.
\item Fracture toughness of TIP4P/ICE+OPLS+UAM under tensile strain.
\item Brittleness of TIP4P/ICE+OPLS+UAM under rapid loading.
\item Separation of two distinct stages of fracture propagation in TIP4P/ICE+OPLS+UAM. 
\end{itemize}


\section{What is the master thesis?}
This thesis is a walkthrough of how to do molecular simulations, how to apply it to methane hydrates, and how to verify and interpret the results. In part, it is also a documentation of my work during the last year. If I have been strugling with a topic, and spent a lot of time figuring it out, this will be reflected in the thesis. These parts will be more elaborate than other parts of the thesis, regardless the scientific value of their contents. An example is the section ``proof-of-concept runs'' in the chapter about modeled systems.  This section represents a lot of work to find the right parameters to go into the model, but the results are not that interesting.

\section{Structure of this thesis}

Chapter \ref{ch:state_of_the_science} sums up relevant aspects of the state of the science of methane hydrates, both experimentally and theoretically, and ends with questions that my research aim to answer.
In part I is also a presentation of relevant parts of the theory of elastricity, fracture mehanics and molecular dynamics. I have sought to explain all theory needed to in principle implement a molecular dynamics program that would do the same job as the one I use. That means I aim to present an overview of what is important for the physics of the system, but not implementation details that are unimportant to the actual outcome of my simulations. Some topics can be explained in detail, while some are too technical, requiring too many details to be interesting in a master thesis. I will refer to research papers that can supplement the presentation where my presentation is incomplete. Part II is the main part of this thesis. Chapter \ref{ch:models} contains the choice of mathematical model for the methane hydrates in addition to a description of the system that I will study. Chapters \ref{ch:tools} and \ref{ch:verification} describes the numerical tools I use and the tools I have developed, along with a brief verification of the numerics. Chapter \ref{ch:modeledsystems} contains the main results. This chapter is mostly organized chronologically after when I did what, and it can be seen as a description of the research process that I have been going through during the last months. Some tools were devoloped during the research process, and these tools are described in detail in chapter \ref{ch:tools}, but it should be clear from the discussion in \ref{ch:modeledsystems} when they were actually introduced. In chapter \ref{ch:summary_conclusions} i summarize my results, conclude base on those results and propose possible topics for future work.


